A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-31 text/abstract.Rmd
incorporate Tim's suggestions for abstract

Cancer affects cellular metabolism, biochemical processes, and the
microenvironment. However, despite some unifying principles, molecular
signatures remain largely indistinguishable from tissue-of-origin, presenting a
major barrier for precision health and individualized medicine. To address this,
we assign tumors into ten clusters based on pathways largely free of bias from
tissue-of-origin, mutational burden or stage. Germline risk genes are equally
distributed among clusters, contrary to the expectation that they determine
molecular subtype. We show that clusters vary in prognosis, modeling risk of
death by cancer type and cluster. We investigate metastasis, and show that most
differences are cluster-specific and complement other clusters, suggesting
convergence. Our observations suggest that driver genes combine with uncommon
genes in tumor-promoting pathways to produce distinct molecular phenotypes.
Thus, we present a coherent view of global tumor biology, and explain how common
cellular dysfunction might arise from tissue-specific mutations.

>>>>

Cancer is an organism-level disease, impacting processes from cellular
metabolism and the microenvironment to systemic immune response. Nevertheless,
efforts to distinguish overarching mutational processes from interactions with
the cell of origin for a tumor have seen limited success, presenting a barrier
to individualized medicine. Here we present a novel, pathway-centric approach,
extracting somatic mutational profiles within and between tissues, largely
orthogonal to cell of origin, mutational burden, or stage. Known predisposition
variants are equally distributed among clusters, and largely independent of
molecular subtype. Prognosis and risk of death vary jointly by cancer type and
cluster. Analysis of metastatic tumors reveals that differences are largely
cluster-specific and complementary, implicating convergent mechanisms that
combine familiar driver genes with diverse low-frequency lesions in
tumor-promoting pathways, ultimately producing distinct molecular phenotypes.
The results shed new light on the interplay between organism-level dysfunction
and tissue-specific lesions.
